\Large \bf Cyclone: Warp Speed Replication for Key Value Stores

### 2017-07-24 paper/paper.tex
Updates.

Finally, we turn our attention to the persistence step in
Figure~\ref{fig:control_plane}. RAFT requires that the log entry be persisted
before it is multicast out to follower replicas. DPDK userspace NIC drivers
operate in DDIO mode~\cite{ddio} where the packet is directly written into the
CPU cache rather than first being DMAed into DRAM and then fetched by the CPU on
demand. Since the NVM is directly attached we need to persist the packet to it
by executing a cacheline flush ({\tt clflush}) instruction for every cacheline
in the packet and the pointer in the pointer buffer to persist these via the
memory bus. This is not too onerous a burden because we can use the newly
introduced {\tt clflush-opt}~\cite{clflush_opt} instruction specifically
intended to efficiently flush to persistent memory without the overhead of the
serialization normally introduced by {\tt clflush}. This allows us to hit full
memory bandwidth on present generation platforms, a quantity in excess of 200
Gb/s per core, well above the near term speeds of network interface cards. We
execute a single serializing {\tt sfence} before the sequence of {\tt clflushes}
to make sure any dirty cacheline due to header manipulation related writes from
the CPU are sent to cache.

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.35]{figures2/race.pdf}
  \caption{Race with multiple physical logs}
  \label{fig:race}
\end{figure}

\section{Horizontal Scaling}
\label{sec:horizontal}
The basic principle in Cyclone is to turn log replication into a software packet
switching problem. A key ingredient in software packet switching is applying
multiple CPU cores to parallelize the control plane, scaling message processing
throughput. In this section, we describe how we can horizontally scale Cyclone's
single \emph{logical} log across multiple \emph{physical} logs. For each
physical log, we run an independent version of RAFT with the corresponding
control plane mapped to a dedicated CPU core. Further, we assume a concurrent
key value store where a set of CPU cores independently run operations from the
physical logs on the shared memory key-value store data structure, synchronizing
among themselves as necessary.

We begin by showing how we divide CPU cores and NIC resources on the plaform across
replication and key value store work, and show how a KV store request flows
across them.  We then describe how a key-value store API can be mapped to
independent physical logs without divergence in replica state. Finally, we
describe special handling in Cyclone for multi-key operations.

\subsection{Core Allocation}
A key-value store integrated with Cyclone runs two types of threads. The first
type of thread does key-value store work. This includes servicing lookup and
update requests and background threads that flush in-memory data structures to
secondary storage and do maintainence work on secondary storage data structures
such as compacting log-structured merge trees~\cite{lsmtree}. The second type of
thread is a dedicated thread per-physical log to run the corresponding control
plane including the associated RAFT state machine. We use static assignment of
threads to cores as recommended for packet switching libraries such as DPDK for
the best possible performance. Therefore we use the terms thread and core
interchangeably in this paper. Client requests are always received by the
control plane core for the physical log the request is directed to. A response
for the core is always sent by the KV store core that finally handles the
request.

An important resource for Cyclone is the NIC that must be shared among the
control plane cores that receive and replicate requests and the KV store cores
that send back responses. Most NICs provide independent transmit/receive queue
pairs and we aim for synchronization free access to these to aid horizontal
scalability with respect to the network resource. DPDK already provides fairly
scalable access to NIC queues, with the queue number being a parameter to send
and receive calls. All that remains for us is to map the queues to the control
plane and KV store cores, keeping in mind that there may be multiple NICs on the
system. We do this by iterating over all control plane cores followed by all KV
store cores. We assign two queue pairs to each control plane core, one to
receive requests from the client and the other to transmit and receive
replication protocol related messages. We assign one queue pair to each KV store
core to send back responses to clients. We are careful to spread the different
queue pairs for a core across the available NICs as far as possible. This is
important due to the asymmetry between the different types of traffic. Queue
pairs carrying replication traffic will likely be more busy than those carrying
client request responses due to the amplification in message count to multicast
the same message to different replicas.

The flow for an update operation in Cyclone is as follows. The network packet is
received by the control plane for the targeted physical log. The packet is then
immediately sent out for replication as described in the previous section. When
responses from a majority quorum is received, the request is passed to the
targeted KV store core for execution. The KV store core processes the request
and sends back the response to the client. We note that replication is pipelined
and decoupled from execution and thus multiple updates are usually in flight on
the network for replication at the same time.

The flow for a read operation is simpler. It is still targeted to a physical log
but instead of replicating it, the control plane core simply passes it on
straight away to the KV store core for execution and response.

The control plane cores communicate with the KV store cores through a one-way
lock-free FIFO queue passing pointers to packets. We leverage the convenient
ring library in DPDK for this.

Finally, an asynchronous source of work in the system is maintenance of the
two-level log. We depart from a 1:1 mapping between the first level NVM log and
the second level flashlog here. Each physical log consists of one NVM log
managed by the corresponding control plane core and multiple flashlogs each of
which is managed as an additional responsibility by a KV store core. We
therefore enforce a many to one mapping between key value store cores and a
front end control plane core - thereby partitioning KV store cores across
control plane cores. A request picks a control plane core and any one of the
associated KV store cores for execution. The request is logged in the NVM log of
the control plane core and garbage collected once it is drained to the flashlog
of the selected KV store core.

\subsection{Leveraging API Commutativity}
Update requests to a key-value store include a key and an associated operation -
such as ``put'' that sets the value for the key or ``merge'' that calls a
user-defined operator to merge a new value for the key into the existing value.
A naive approach that maps a key to a random physical log and subsequently to a
random core on the system for execution runs into the determinism problem -
replicas could diverge with respect to the final value assigned to a key on
different replicas. We illustrate this problem with an example in
Figure~\ref{fig:race} where two updates to the same key race with each other on
different physical logs and execution cores to end up being applied in different
orders on different replicas, leading to a divergence of state.

The solution to this problem is observe that the key-value store is a concurrent
data structure that ultimately admits a serialization of operations to
it. Further, consecutive operations to different keys in the serial history can
be interchanged, because operations to different keys are
\emph{commutative}. The correct way to use different physical logs is therefore
to simply hash the key to select the physical log \emph{and} the core on which
the operation executes. This guarantees determinism as the serial history of
two different replicas can be transformed into the same serial history by
reordering operations to different keys. Operations to the same key are already
in the same order in the serial histories by virtue of being replicated on the
same physical log and being executed on the same KV store core.

Further relaxations to the mapping scheme are possible when operations to the
same key are also commutative. A good example is when key value stores are used
to maintain counters. Different increment operations to the same counter are
commutative if one is prepared to tolerate non-determinism in the sequence of
updates seen by a reader with the counter reaching the same eventual value. This
specific use-case, in fact, lead to the addition of the merge operation in
Rocksdb, a key value store that we use for evaluation. Merge operations
therefore could be dispatched to a random physical log with no harm. For this
paper however, we stick to the stricter case that operations to the same
key are not commutative. However, Cyclone's client side libraries leave it to
the user to specify the exact physical log and key-value store core to use in
order to enable more relaxed mappings when necessary.

We note that commutativity as assumed by Cyclone is an \emph{API level
  property}. There is no assumption as to how application cores synchronize to
access the shared key-value state or the underlying data structures for holding
key-value pairs. We emphasize that this means replicas ultimately have the same
content but not necessarily the same memory state.

Finally, Cyclone provides seraializable reads but does not - at the moment -
guarantee the ability to read one's own writes, repeatable reads or more
generally, linearizable reads. The reason is that on a failover the new leader
replica might still have update operations in its NVM logs that are yet to be
applied to the key value store. However the original failed leader replica might
have applied these updates and a client might have already observed them. Users
who desire linearizable reads can issue a special API call to ensure that reads
are also routed through the logs. Flushing the NVM logs on failover is a better
solution that we are currently in the process of engineering into Cyclone.

\subsection{Ganged Operations}
Key value stores (such as Rocksdb) usually provide a batched write operation
that atomically updates a set of key-value pairs. A batched write is often a key
primitive for building more sophisticated operations such as
transactions. Batched writes require special handling in Cyclone due to the
requirement that we route a write request by hashing the key to select both the
control plane core (physical log) as well as KV store core. A batched write has
multiple keys and therefore we need to synchronize the progress of the operation
across multiple physical logs (and associated replication quorums) as well as KV
store cores. In effect, the serial history contains a grouped set of operation
on multiple keys as a single atomic operation. All serial histories across
replicas therefore continue to map to the same equivalent serial history modulo
reordering of operations to different keys.

One way to do this is to use a distributed transaction, using
two-phase commit to couple the otherwise independent replication quorums and KV
store cores. However this method has a number of problems. First, it introduces
an additional network round trip that is normally not necessary for a key value
store with a single shared memory image holding all the keys. Second, handling
failures with distributed transactions would have caused an explosion in
complexity in Cyclone, something we wished to avoid in the interests of
practical deployment to independent end users who need to understand and 
maintain Cyclone as part of their embedded KV store infrastructure.

Our solution instead is to use a technique we term ganged operations. It is
based on the observation that in the failure free case, executing a batched
write across multiple control plane cores and KV store cores does not look very
much different from a synchronous shared memory barrier. The challenge is to
effectively manage failures in replication without causing participating cores
to stop making progress. The first step is to reconfigure RAFTs leader election
protocol to ensure that the leaders for all the physical logs are co-located on
the same host. Clients always dispatch batched writes to a fixed control plane
core (called the injection core) which is then responsible for forwarding the
request to the participating control plane cores (for corresponding physical
logs). The injection core uses DPDK's packet cloning primitives to avoid
dataplane work while forwarding copies of the same batched write packet to
multiple control plane cores. The injection core also adds a ``nonce'' - a
unique timestamp to the packet. In addition, the injection core adds a unique
view number to the packet containing the term numbers of all the participating
RAFT instances (read from shared memory of the co-located leaders).  The flow
for ganged operations is described in the pseudocode of
Figure~\ref{fig:ganged_ops}. We defer the discussion of how we generate the
nonce to later in this section. We also assume that a unique barrier is
allocated in shared memory for each ganged operation. We discuss later how this
is done without using dynamic allocation.

In the absence of failures Figure~\ref{fig:ganged_ops} is straightforward. Once
replication is complete on all participating physical logs, the corresponding KV
store cores execute \\{\tt event\_replicated\_ganged\_op}. They use a bitmask
for the barrier setting the appropriate bit for themselves. Once all KV store
cores have arrived at the barrier the operation is executed. A distinguished
leader core (we use the minumum numbered participating KV store core) is
responsible for sending back the response to the client. Two types of failure
need to be considered. First, replication on the physical log corresponding to
one of the non-leader cores could fail. The remaning cores would never progress
past the barrier. To solve this the leader core monitors the remaning cores for
their current view - which is the RAFT term of the last executed log entry. If
the view has moved past the view in the packet, replication failed on that
physical log. The leader then sets the bit for the core that will no longer
participate in the barrier to aid progress (a technique we borrowed from
lock-free data structures~\cite{lock-free}). It also sets a failed flag to
signal failure to the other participating cores. The second failure case is that
the leader core might not receive the operation if it fails during replication
from corresponding physical log. This case is detected in a similar manner by
the remaining cores by monitoring the published view from the leader core in
shared memory. We add that on a leadership change due to timeout we enqueue and
execute a nop on the physical log and all associated KV store cores to ensure
previous entries are committed - a corner case in the RAFT protocol.

\begin{figure*}
  \centering
\begin{verbatim}
    // Control plane core
    event_recv_ganged_op(packet)
    {
      Begin replication of packet
    }

    // KV store core
    event_replicated_ganged_op (packet, barrier)
    {
      if(leader KV core)
        atomic set bit me in barrier.mask         
        do
         for each KV store core in packet
          if core_public_data[core].view > packet.view
           barrier.failed = true 
           atomic set bit for core in barrier.mask
        while barrrer.mask != mask of participating KV cores
        if barrier.failed
         Ganged replication failed. Send retry to client. return.
        else       
         execute operation
         send response
      else
        wait until 
         core_public_data[leader core].view > packet.view OR
         barrier.mask == mask of participating KV cores
        if core_public_data[leader core].view > packet.view OR
           barrier.failed
         Ganged replication failed. Send retry to client. return.
        else
         execute operation
    }
\end{verbatim}
\caption{Ganged Operation}
\label{fig:ganged_ops}
\end{figure*}

The assumption in Figure~\ref{fig:ganged_ops} is that each ganged operation is
mapped to a unique barrier. We achieve this by using a fixed piece of memory
owned by the leader core to hold the barrier and write the nonce to it in order
to indicate that the barrer is active for the corresponding batched write
operation. Non-leader cores watch for the nonce to know when to execute the
ganged operation barrier in Figure~\ref{fig:ganged_ops}, while also monitoring
the leader's published view to detect the case where the ganged operation fails
to replicate on the leader's physical log.

Finally, we describe how we generate the nonce. The nonce is generated on the
injector core by concatenating the ethernet MAC ID of the first NIC on the
system with a 64 bit value that is the number of CPU timestamp counter cycles
since epoch time (read from the real time clock at startup plus the number of
cycles from the CPU {\tt rdtsc} instruction). The nonce can only be repeated if
the same machine manages to fail and come back up in less time than the real
time clock drift (controlled with NTP), a possibility that we discount.

Ganged operations possibly consititute the most complex part of Cyclone but the
code weighs in at well under a couple of hundred lines. We believe that
this additional complexity is still small compared to distributed transactions.

>>>>

We now turn our attention to the persistence step in
Figure~\ref{fig:control_plane}. RAFT requires that the log entry be persisted
before it is multicast out to follower replicas. DPDK userspace NIC drivers
operate in DDIO mode~\cite{ddio} where the packet is directly written into the
CPU cache rather than first being DMAed into DRAM and then fetched by the CPU on
demand. Since the NVM is directly attached we need to persist the packet to it
by executing a cacheline flush ({\tt clflush}) instruction for every cacheline
in the packet and the pointer in the pointer buffer to persist these via the
memory bus. This is not too onerous a burden because we can use the newly
introduced {\tt clflush-opt}~\cite{clflush_opt} instruction specifically
intended to efficiently flush to persistent memory without the overhead of the
serialization normally introduced by {\tt clflush}. This allows us to hit full
memory bandwidth on present generation platforms, a quantity in excess of 200
Gb/s per core, well above the near term speeds of network interface cards. We
execute a single serializing {\tt sfence} before the sequence of {\tt clflushes}
to make sure any dirty cacheline due to header manipulation related writes from
the CPU are sent to cache.

@@

%\begin{figure}
%  \centering
%  \includegraphics[scale=0.35]{figures2/race.pdf}
%  \caption{Race with multiple physical logs}
%  \label{fig:race}
%\end{figure}

Althought the work described in the previous section significantly boosts the
network performance of replication in Cyclone, it still does not come close to
saturating the capabilities of even the commodity network interface card that we
use. This is because NICs today encapsulate significant parallelism in terms of
multiple send and receive queue pairs. Exploiting this capability to improve
throughput requires us to remove the bottleneck of a single sequential log in
RAFT. To this end, we extend the implementation described thus far to run
multiple copies of the RAFT consensus protocol each maintaining and replicating
its own two level log. We now refer to these logs as physical logs. All these
instances however exist in the same process address space as the key value store
application itself and therefore manipulate the same application.

The first key question we deal with is - how do Cyclone clients decide which
physical log to send a {\tt PUT} or {\tt GET} request to ? A key constraint here
is that all replicas should agree on the seqeunce of updates to a key - as
otherwise they would diverge in terms of state. Allowing updates to the same key
to go to different physical logs would lead to a divergence as there is no
guarantee about the order in which the independent RAFT instances would apply
the update. Therefore all updates to a key must go to the same physcial
log. A similar constraint applies to reads in order to ensure that any committed
writes are seen by readers. Allowing reads to a key to go to a different RAFT
instance than the write means that it could be served on a machine where the
write is yet to be applied but where the machine is the leader for the RAFT
instance. All reads and write to a key must therefore go to the same physical
log. We achieve this by hashing the key to select the physical log.

It turns out that this simple \emph{necessary} scheme is also \emph{sufficient}
to ensure linearizability in the face of failure. Clearly the system is
linearizable for a single key since it involves only a single physical log. In
addition, one can switch operations to different keys in a history without
affecting the correctness of the history because operations to different keys
are commutative. Any history of the system is therefore linearizable.

Multiple physical logs in Cyclone operate in a shared nothing manner by
partitioning the NVM and SSD space evenly between them and by allocating
dedicated NIC queue pairs to each instance of RAFT. This works well because the
memory hierarchy including the NVM and the SSD efficiently support concurrent
operations. The only synchronization necessary is when doing reads or writes to
the single shared key value stores.  The level of concurrency therefore is
constrained only by the concurrency available in the software architecture of
the persistent key value store itself, for which a number of good designs
already exist~\cite{flodb}.

In our discussion thus far, we have focused on operations to a single
key. Supporting atomic writes or reads to multiple keys while maintaining
linearizability requires special handling with multiple physical logs so as not
to break linearizability - that we discuss next.

\subsection{Ganged Operations}
Supporting reads or writes to multiple keys that map to different physical logs
requires a ganged operation across those logs such that all the operations
complete together in the face of failures. Since our starting point is a single
key value store, we constrained Cyclone's design early on to avoid the need for
any two phase commit.

The first key problem we need to deal with is that the leaders for the different
RAFT instances can be located on different machines. To avoid having a client
co-ordinate different machines for a single operation, we constrain the RAFT
leader election algorithm for the different RAFT instances to converge to the
same machine. We do this by triggering re-elections for a RAFT instance other
than that of the first physical log until it is on the same machine as the
\emph{first} physical log. This process is therefore resilient to failed
machines. RAFT requires that the leader have the most up to date log. To ensure
that the process converges we therefore ensure that the currently elected leader
brings a majority quroum up to date before triggering a relection. We also do
not accept any requests from clients until all leaders are co-located. This
process is guaranteed to converge as long as all no network partitions occur
\emph{within a machine} i.e. some RAFT instances are able to send and receive
packets, while others are not, a possiblity we discount due to our single
process design.

Next, we need to simultaneously inject the ganged operation into all
participating physical logs on the machine hosting the leaders. Clients always
dispatch batched writes to a fixed co-ordinating physical log/RAFT instance
(henceforth called the co-ordinator), which is then responsible for forwarding
the request to the participating logs. We make use of packet cloning primitives
to avoid making a physical copy of the packet, generating indirect references
instead. The co-ordinator also adds a ``nonce'' - a unique timestamp to the
packet. In addition, the co-ordinator adds a unique view number to the packet
containing the term numbers of all the participating RAFT instances (read from
shared memory of the co-located leaders).  The event that applies a ganged
operation is described in the pseudocode of Figure~\ref{fig:ganged_ops}. We
defer the discussion of how we generate the nonce to later in this section. We
also assume that a unique barrier is allocated in shared memory for each ganged
operation. We discuss later how this is done without using dynamic allocation.

The complexity in Figure~\ref{fig:ganged_ops} arises from the need to handle
both shared memory concurrency and failure during replication. Without failure,
Figure~\ref{fig:ganged_ops} is straightforward. A barrier is executed on all
RAFT instances. Once replication is complete on all participating physical logs
(or this is a read ganged operation), a state indicated by all necessary bits
being set in the barrier mask, the RAFT instances simultaneously execute the
ganged operation.

Each RAFT instance always publishes the current term it is operating in. A
failure causes a new leader to be elected moving forward the term. If the
participants of a ganged operatation do not detect this case, they could be left
waiting for ever as ganged operation might never successfully replicate in the
failed over RAFT instance. We detect failure on any RAFT instance by having each
instance publish its current term and continuously comparing the view in the
ganged operation to it. If any RAFT instance has moved past that term, the
ganged operation is then terminated by having one of the still alive participants set a
failure flag and write the appropriate bit in the barrier mask on behalf of the
failed participant, moving execution past the barrier wait - a technique we
borrow from lock-free data structures~\cite{lock-free}.

\begin{figure*}
  \centering
\begin{verbatim}
    // Apply ganged operation
    event_apply_ganged_op (packet, barrier)
    {
      if(co-ordinator)
        atomic set bit me in barrier.mask         
        do
         for each participant p in operation
          if public_data[p].view > packet.view
           barrier.failed = true 
           atomic set bit for p in barrier.mask
        while barrrer.mask != mask of all participants
        if barrier.failed
         Ganged replication failed. Send retry to client. return.
        else       
         execute operation
         send response
      else
        wait until 
         public_data[co-ordinator].view > packet.view OR
         barrier.mask == mask of participating KV cores
        if public_data[co-ordinator].view > packet.view OR
           barrier.failed
         Ganged replication failed. Send retry to client. return.
        else
         execute operation
    }
\end{verbatim}
\caption{Ganged Operation}
\label{fig:ganged_ops}
\end{figure*}

The assumption in Figure~\ref{fig:ganged_ops} is that each ganged operation is
mapped to a unique barrier. We achieve this by using a fixed piece of memory
owned by the co-ordinator to hold the barrier and write the nonce to it in order
to indicate that the barrer is active for the corresponding ganged
operation. Participants watch for the nonce to know when to execute the
ganged operation barrier in Figure~\ref{fig:ganged_ops}, while also monitoring
the leader's published view to detect the case where the ganged operation fails
to replicate on the co-ordinator's physical log.

Finally, we describe how we generate the nonce. The nonce is generated on the
co-ordinator RAFT instance by concatenating the ethernet MAC ID of the first NIC
on the system with a 64 bit value that is the number of CPU timestamp counter
cycles since epoch time (read from the real time clock at startup plus the
number of cycles from the CPU {\tt rdtsc} instruction). The nonce can only be
repeated if the same machine manages to fail and come back up in less time than
the real time clock drift (controlled with NTP), a possibility that we discount.

Ganged operations possibly consititute the most complex part of Cyclone but the
code weighs in at well under a couple of hundred lines. We believe that
this additional complexity is still small compared to distributed transactions
using two-phase commit that also need to recover from failure used in systems
such as FARM~\cite{farm}.
